Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-31 _tex/index.tex
Built site for gh-pages

Over the past decade, numerous studies have documented the effectiveness
of the \emph{comparative judgment} (CJ) method \citep{Thurstone_1927}
for assessing competencies and traits. These studies have evaluated CJ
from three main perspectives: its ability to produce reliable and valid
trait scores, its practical applicability, and its time efficiency.
Research on reliability and validity shows that CJ can generate precise
and consistent scores
\citep{Pollitt_2012a, Pollitt_2012b, Coertjens_et_al_2017, Goossens_et_al_2018, Verhavert_et_al_2019, Crompvoets_et_al_2022, Bouwer_et_al_2023}
that accurately represent the traits being measured
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bouwer_et_al_2023}.
Research on practical applicability highlights CJ's versatility across
both educational and non-educational contexts, presenting it as an
efficient and effective alternative for measurement and evaluation
\citep{Pollitt_2004, Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.
Lastly, research on time efficiency suggests that CJ can offer at least
equal, if not significant, time savings when evaluating stimuli compared
to traditional marking methods
\citep{Pollitt_2012a, Pollitt_2012b, Coertjens_et_al_2017, Goossens_et_al_2018}.

Nevertheless, despite the growing number of studies on CJ, unsystematic
and fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's structural and measurement model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role and impact of comparison algorithms on the method's reliability and
validity. The following sections will discuss each of these issues in
detail, followed by the introduction of a theoretical model and its
statistical translation, which aim to address all three concerns
simultaneously.

@@

In a typical CJ study, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
\citep{Andrich_1978, Bramley_2008}. This model specifies how latent
variables are estimated from manifest variables
\citep{Everitt_et_al_2010}. In CJ, multiple judges engage in several
rounds of pairwise comparisons to assess the relative manifestation of a
trait between two stimuli. Each comparison generates a dichotomous
outcome, indicating which stimulus is perceived to exhibit a higher
degree of the trait. The BTL model then uses these observed outcomes to
estimate scores that represent the latent trait of interest
\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.

Moreover, researchers often use these BTL-generated scores or their
transformations separately, conducting further analyses or hypothesis
testing. The CJ literature shows how these scores have been employed to
identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores for separate analyses and tests. A key consideration is that
BTL-generated scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty when conducting separate analyses
and tests can inflate their precision and statistical power. This, in
turn, increases the risk of committing a type I error
\citep{McElreath_2020}, which is when a null hypothesis is incorrectly
rejected \citep{Everitt_et_al_2010}. To mitigate this risk, principles
from Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response Theory
(IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017} recommend
conducting these analyses and tests within a structural model that
accounts for both the scores and their uncertainties, rather than
treating them separately. Thus, an integrated approach combining CJ's
structural and measurement models can offer significant advantages.

@@

\textbf{Authors' contributions:} \emph{Conceptualization:} S.G., S.DM.,
T.vD., and J.M.R.E; \emph{Methodology:} S.DM., T.vD., and J.M.R.E;
\emph{Software:} J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal
Analysis:} J.M.R.E.; \emph{Investigation:} J.M.R.E; \emph{Resources:}
S.G., S.DM., and T.vD.; \emph{Data curation:} J.M.R.E.; \emph{Writing -
original draft:} J.M.R.E.; \emph{Writing - review \& editing:} S.G.,
S.DM., and T.vD.; \emph{Visualization:} J.M.R.E.; \emph{Supervision:}
S.G. and S.DM.; \emph{Project administration:} S.G. and S.DM.;
\emph{Funding acquisition:} S.G. and S.DM.

>>>>

{In \emph{comparative judgment} (CJ) studies, judges assess the presence
of a trait or competence by conducting pairwise comparisons of stimuli
\citep{Thurstone_1927, Pollitt_2004, Pollitt_2012a}. The comparison
produces a dichotomous outcome, indicating which stimulus is perceived
to possess a higher level of the trait. After conducting multiple rounds
of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL)
model \citep{Bradley_et_al_1952, Luce_1959} to process the outcomes and
estimate scores that reflect the underlying trait of interest. CJ has
been successfully employed in assessing the quality of written texts
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.}

Numerous studies have documented the effectiveness of CJ in assessing
various traits and competencies over the past decade. These studies have
emphasized three aspects of the method's effectiveness: its reliability,
validity, and practical applicability. Research on reliability indicates
that CJ requires a relatively small number of pairwise comparisons
\citep{Verhavert_et_al_2019, Crompvoets_et_al_2022} to produce trait
scores that are as precise and consistent as those generated by other
assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
Furthermore, evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}. On
the other hand, research on validity suggests that scores generated by
CJ can accurately represent the traits being measured
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Finally, research on practical applicability highlights the method's
versatility across both educational and non-educational contexts
\citep{Jones_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the growing number of studies on CJ, unsystematic
and fragmented research approaches in the literature have left several
critical issues unaddressed. This research primarily focuses on three:
the apparent disconnect between CJ's measurement and structural model,
the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in CJ's measurement model, and the unclear
role of comparison algorithms on the method's reliability and validity.
The following sections will discuss each of these issues in detail,
followed by the introduction of a theoretical model and its statistical
translation, which aim to address all three concerns simultaneously.

@@

In a typical CJ study, the BTL model serves as the measurement model for
CJ \citep{Andrich_1978, Bramley_2008}. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
the BTL model uses the dichotomous outcomes resulting from the pairwise
comparisons (the manifest variables) to estimate scores that reflect the
underlying quality level of the texts (the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

Researchers then typically use the estimated BTL scores, or their
transformations, to conduct additional analyses or hypothesis testing.
The literature indicates that these scores have been used to identify
`misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the latent trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature cautions against using estimated
scores to conduct additional analyses or tests. A key consideration is
that BTL scores are parameter estimates that inherently carry
uncertainty. Ignoring this uncertainty when conducting additional
analyses and tests can inflate their precision and statistical power,
increasing the risk of committing a type I error \citep{McElreath_2020},
which is when a null hypothesis is incorrectly rejected
\citep{Everitt_et_al_2010}.

To mitigate these risks, principles from Structural Equation Modeling
(SEM) \citep{Hoyle_et_al_2023, Kline_et_al_2023} and Item Response
Theory (IRT) \citep{deAyala_2009, Fox_2010, vanderLinden_et_al_2017}
recommend conducting these analyses and tests within a structural model
that accounts for both the scores and their uncertainties, rather than
treating them separately. Thus, an integrated approach combining CJ's
structural and measurement models can offer significant advantages.

@@

\textbf{AI-assisted technologies in the writing process:} The authors
used ChatGPT, an AI language model, during the preparation of this work.
They occasionally employed the tool to refine phrasing and optimize
wording, ensuring appropriate language use and enhancing the
manuscript's clarity and coherence. The authors take full responsibility
for the final content of the publication.

\textbf{CRediT authorship contribution statement:}
\emph{Conceptualization:} S.G., S.DM., T.vD., and J.M.R.E;
\emph{Methodology:} S.DM., T.vD., and J.M.R.E; \emph{Software:}
J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal Analysis:} J.M.R.E.;
\emph{Investigation:} J.M.R.E; \emph{Resources:} S.G., S.DM., and T.vD.;
\emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
J.M.R.E.; \emph{Writing - review \& editing:} S.G., S.DM., and T.vD.;
\emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;
\emph{Project administration:} S.G. and S.DM.; \emph{Funding
acquisition:} S.G. and S.DM.
